Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-01-07 MANUSCRIPT.qmd
Abe Edits Draft

---
title: Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands | Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records | Estimating Base-Flow Index in Gauged and Ungauged Dryland Basins
author:
  - name: Caelum Mroczek
    orcid: 0009-0005-6695-0134
    corresponding: true
    email: csm428@nau.edu
    roles:
      - Conceptualization
      - Investigation
      - Methodology
      - Formal analysis
      - Visualization
      - Software
      - Data curation
      - Writing - Original Draft
    affiliations:
      - 'School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA'

@@

Advancements in machine learning provide tools to predict hydrologic indices in ungauged basins, addressing the limitations of sparse streamgage networks. To tackle the challenge of quantifying base flow in ungauged catchments, numerous studies have applied both regression and machine learning methods. [@ahiablame2013] found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate. @beck2013 overcame the nonlinearity of basin characteristics and improved results of multivariate analyses by using artificial neural networks to estimate BFI globally. @singh2018 implemented a random forest algorithm to predict long-term BFI for ungauged catchments across New Zealand. These applications demonstrate the versatility and effectiveness of machine learning in capturing complex ecohydrologic dynamics and improving our understanding of groundwater contributions to streamflow.

This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI. Regional trends in base flow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, evapotranspiration (ET~O~​), and temperature over the same periods. Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020. This approach helps address the spatial gaps in the streamgage network, which is relatively sparse across the state. The results offer novel insights into low-flow processes in both gauged and ungauged basins, enhancing our understanding of climate controls on consistent flows in Arizona. This study contributes to a more comprehensive view of hydrological dynamics in the context of arid and semi-arid landscapes.

@@

Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study. Streamgages were selected depending on criteria to ensure the applicability of each site. Following the findings of @odonnell2016, which determined that 8–10 years of calibration data are necessary to account for climate variability in paired watershed studies in the region, a minimum record length of 10 years was required. Additionally, years with more than 30 missing days of streamflow data were excluded from the analysis. This study focuses on natural, streamflow-influencing dynamics, so streamgages affected by regulation or diversions were excluded. These streamgages were identified through annual reports on water data published by the USGS [@usgs2010]. Furthermore, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact. After applying these selection criteria, 205 USGS streamgages with acceptable periods of record were included in the study (@fig-study-area). Periods of record ranged from 10 to 112 years, with a median of 28 years.

While our data selection ensures a robust record for this analysis, it also highlights gaps in spatial coverage that the machine learning model aims to address. Arizona has 184 active USGS streamflow stations (as of 2024) across an area of 295,253 km², whereas Indiana has 189 active stations within an area of 94,326 km². This translates to a streamgage density of approximately 2.004 gages per 1,000 km² in Indiana, over three times higher than Arizona's density of 0.623 gages per 1,000 km². This stark difference underscores the relative sparsity of Arizona's streamgage network, particularly in the context of its larger geographic area and the unique hydrological challenges posed by its arid and semi-arid landscapes.

@@

A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects. Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins. By calculating these percentages, we aimed to get a more comprehensive understanding of landscape composition across space. Land cover from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins. Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.

@@

DATA LIMITATIONS - ABE

>>>>

---
title: Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records
subtitle: Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands |  | Estimating Base-Flow Index in Gauged and Ungauged Dryland Basins
author:
  - name: Caelum Mroczek
    orcid: 0009-0005-6695-0134
    corresponding: true
    email: csm428@nau.edu
    roles:
      - Conceptualization
      - Investigation
      - Methodology
      - Formal analysis
      - Visualization
      - Software
      - Data curation
      - Writing - Original Draft
    affiliations:
      - 'School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA'

@@

Advancements in machine learning provide tools to predict hydrologic indices in ungauged basins, addressing the limitations of sparse streamgage networks. To tackle the challenge of quantifying base flow in ungauged catchments, numerous studies have applied both regression and machine learning methods. @ahiablame2013 found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate. @beck2013 overcame the nonlinearity of basin characteristics and improved results of multivariate analyses by using artificial neural networks to estimate BFI globally. @singh2018 implemented a random forest algorithm to predict long-term BFI for ungauged catchments across New Zealand. These applications demonstrate the versatility and effectiveness of machine learning in capturing complex ecohydrologic dynamics and improving our understanding of groundwater contributions to streamflow.

This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI. Regional trends in base flow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, evapotranspiration (ET~O~), and temperature over the same periods. Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020. This approach helps address the spatial gaps in the streamgage network, which is relatively sparse across the state. The results offer novel insights into low-flow processes in both gauged and ungauged basins, enhancing our understanding of climate controls on consistent flows in Arizona. This study contributes to a more comprehensive view of hydrological dynamics in the context of arid and semi-arid landscapes.

@@

Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study. Streamgages were selected depending on criteria to ensure the applicability of each site. Following the findings of @odonnell2016, which determined that 8–10 years of calibration data are necessary to account for climate variability in paired watershed studies in the region, a minimum record length of 10 years was required. Additionally, years with more than 30 missing days of streamflow data were excluded from the analysis. This study focuses on natural, streamflow-influencing dynamics, so streamgages affected by regulation or diversions were excluded. The regulated river streamgages were identified through annual reports on water data published by the USGS [@usgs2010]. Furthermore, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact. After applying these selection criteria, 205 USGS streamgages with acceptable periods of record were included in the study (@fig-study-area). Periods of record ranged from 10 to 112 years, with a median of 28 years.

While our data selection ensures a robust record for this analysis, it also highlights gaps in spatial coverage that the machine learning model can address. Arizona has 184 active USGS streamflow stations (as of 2024) across an area of 295,253 km², whereas Indiana (a humid state in the U.S.) has 189 active stations within an area of 94,326 km². This translates to a streamgage density of approximately 2.004 gages per 1,000 km² in Indiana, over three times higher than Arizona's density of 0.623 gages per 1,000 km². This stark difference underscores the relative sparsity of Arizona's streamgage network, particularly in the context of its larger geographic area and the unique hydrological challenges posed by its arid and semi-arid landscapes.

@@

A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects. Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins. By calculating these percentages, we aimed to get a more comprehensive understanding of landscape composition across space. Land cover was acquired from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins. Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.

@@

While there has been an increase of stream gages in Arizona to comply with other imperatives, such as the Clean Water Act, these gages are not of sufficient quality for base-flow analyses. New, non-USGS stream gages are typically installed by flood control districts for flood flow monitoring, and do not accurately track base flow. New USGS stream gages have been added over the past 10 years to address instream flow rights and to improve the density of monitoring in the future.
